---
title: "Basic R Syntax"
author: |
  | "Kumar Rahul
  |
  | kumarrah@iimb.ac.in; rahul235@gmail.com
output:
  bookdown::html_document2:
    toc: TRUE 
    toc_depth: 4
    toc_float: TRUE
    fig_width: 7
    fig_height: 5
    fig_caption: TRUE
    number_sections: TRUE
footer: AwesomeStats Consulting Pvt Ltd
---

### Declaring and assigning a variable

```{r}
a = 5
```

```{r}
class(a)
```

```{r}
a  = "This is a string"
```

```{r}
class(a)
```

## Printing values of a variable

```{r}
print(a)
```

### Conditional Statement

```{r}
a = 5

if (a > 4) {
    print("True")
} else {
    print("False")
}
```

### Range of numbers

```{r}
print(1:10)

seq(1,10,1)

```

```{r}
listOfNums = list(1:10)
typeof(listOfNums)
```

```{r}

Reduce('+',unlist(listOfNums))

```

```{r}
sum = 0
for (i in seq(1,10)){
        sum = sum + listOfNums[[1]][i]
}
```

```{r}
sum
```

```{r}
sum = 0
i = 0
while (i <= 10){
    sum = sum + i
    i = i + 1
}
```

```{r}
sum
```

### Defining Function

```{r}
multiply = function( a, b){
    return (a*b)
}
```

```{r}
multiply(4,5)
```

```{r}
addNumbers = function(a, b){
    return (a + b)
}
```

```{r}
addNumbers(4, 5)
```

```{r}
addNumbers(2.444, 5.6)
```

```{r}
addNumbers(c(1,3), c(4,5))
```

### Parameters optional

```{r}
addNumbers = function(a, b = 4){
    return (a + b)
}
```

```{r}
addNumbers(4)
```

```{r}
c = addNumbers(4, 3)
```

```{r}
c
```


## Collections

### List

* Vector in R is created by using c() or seq().
* The list() in R is used to create a list whose elements can be of any type i.e. numeric, character, matrix, or a list/vector itself.

```{r}
# Empty List
a = list()
```

```{r}
class(a)
```

```{r}
## List initialized
a = list(1,2,3,4,5)
```

```{r}
a[1]
```

```{r}
a[1:4]
```

#### Reversing a list

```{r}
a= rev(a)
```

```{r}
a
```

Create a list called b, which contains square value of each element in a.

```{r}
b = list()
for(i in seq(1, length(a),1)){
    b = (append(b, a[[i]]*a[[i]]))
}
```

```{r}
b
```

## Use of filter, map and reduce

* The Map() function takes in a function and a list as argument. The function is called with a anonymous function and a list. Map returns a new list which contains all the anonymous modified items returned by that function for each item.

Let's use the anonymous function to square the number contained in the list `a`

```{r}
a = seq(1,10)
a
```

```{r}
list_squared = Map(function (x) x^2, a)
print(list_squared)
```

* The Filter() function construct is like the map function but is used filter out the elements of a sequence, for which the function returns True. Here is a small program that returns the odd numbers from an input list:

```{r}
list_odds = Filter(function (x) x%%2!=0, a)
print(list_odds)
```

```{r}
list_filter = Filter(function(x) (x > 4) & (x < 7), a)
list_filter
```

* The reduce() function also takes in a function and a list as argument. This function can also be called with an anonymous function and a list.The output will be a new reduced result. This performs a repetitive operation over the pairs of the list.

```{r}
reduce_sum = Reduce(function (x,y) x+y, a)
reduce_sum
```

### Set

```{r}
a = c(1,2,3,4,3,6,7,3,2)
```

```{r}
a
```

```{r}
d = union(a,a)
```

```{r}
d
```

intersect, setdiff, setequal are other useful functions for set operations in R.

### Strings

```{r}
str1 = "this is a string"
```

```{r}
str2 = 'this is a string'
```
