---
title: "Titanic Exploration"
author: |
  | "Kumar Rahul
  |
  | kumarrah@iimb.ac.in; rahul235@gmail.com
output:
  bookdown::html_document2:
    toc: TRUE 
    toc_depth: 4
    toc_float: TRUE
    fig_width: 7
    fig_height: 5
    fig_caption: TRUE
    number_sections: TRUE
footer: AwesomeStats Consulting Pvt Ltd
---

## Importing library

```{r}
library(dplyr)
library(magrittr)
```

# Case - Titanic Dataset

The data has been taken from : https://www.kaggle.com/c/titanic/data

The data description:


> * PassengerId: type should be integers
* Survived: Survived or Not
* Pclass: Class of Travel
* Name: Name of Passenger
* Sex: Gender
* Age: Age of Passengers
* SibSp: Number of Sibling/Spouse aboard
* Parch: Number of Parent/Child aboard
* Ticket
* Fare
* Cabin
* Embarked: The port in which a passenger has embarked. C - Cherbourg, S - Southampton, Q = Queenstown

We will use the titanic dataset to understand data load and munging using R base

## Read the dataset, which is in csv format

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/Rahul/Documents/Rahul Office/IIMB/Work @ IIMB/Training Material/Concepts/R/R_Programming")

setwd("/Users/Rahul/Documents/Rahul Office/IIMB/Work @ IIMB/Training Material/Concepts/R/R_Programming")
```

```{r}
getwd()
```

```{r}
titanic_df = read.csv('./files/titanic.csv', stringsAsFactors = TRUE)
```

Data type of titanic dataframe
```{r}
class(titanic_df)
```

## Print the first few rows
head() also takes an argument n, which specifies how many records will be printed.

```{r}
head(titanic_df)
```

```{r}
tail(titanic_df,10)
```

## Statistical Summary

Use `summary` method

```{r}
summary(titanic_df)
```

summary of a individual vraibale
```{r}
summary(titanic_df$Survived)
```

Dimensions of the data
```{r}
dim(titanic_df)
```

## Indexing and Slicing

> Select first n rows across all columns

```{r}
titanic_df[,]
```

```{r}
titanic_df[1:5,1:3]
```

> Select first 4 rows and first 3 columns

```{r}
titanic_df[1:4,1:3]
```

> Select first 2 columns of the last row

```{r}
titanic_df[nrow(titanic_df),c(1:2)]
```

## Check the data type

> * Data type
* dimensions.. how many row and columns?
* Structure of the data

Dimensions of data

```{r}
dim(titanic_df)
```

How many total entries? What are the columns and their types. Each column has how many not-null values?

Structure of Data

```{r}
str(titanic_df)
```

Structure of specific column

```{r}
str(titanic_df$Sex)
```

### Select Specific columns and rows¶

> * Select the Survived column and display the first 5 entries
* How many people survived and what is the percentage?

```{r}
titanic_df[c('Survived')][1:10,]
```

```{r}
head(titanic_df[c('Survived')],10)
```

```{r}
class(titanic_df$Survived[0:5])
```

How many passengers survived and how many did not survive
```{r}
table(titanic_df$Survived)
```

The other way
```{r}
table(titanic_df[['Survived']])
```

What percentage of passengers survived
```{r}
prop.table(table(titanic_df$Survived))
```


```{r}
prop.table(table(titanic_df[['Survived']]))
```

## Selecting multiple columns. 

Multiple column names should be provided as a collection

```{r}
head(titanic_df[c('Survived','Age')])

titanic_df[c('Survived','Age')][1:6,]
```

## Cross tabulation of data

> * Pivot gender with survived

```{r}
table(titanic_df$Sex,titanic_df$Survived)
```

Percentage as row sum
```{r}
prop.table(table(titanic_df$Sex,titanic_df$Survived),1)
```

Percentage as column sum
```{r}
prop.table(table(titanic_df$Sex,titanic_df$Survived),2)
```

Percentage as overall sum
```{r}
prop.table(table(titanic_df$Sex,titanic_df$Survived))
```

## Condition based filters

* How many children below 5 years age were on board the ship?
```{r}
new_df = titanic_df[ which(titanic_df$Age <=5), ]
dim(new_df)[1]
```

* How many children survived who are less than 5 years old?
```{r}
table(titanic_df[ which(titanic_df$Age <= 5), ]$Survived)
```

```{r}
dim(titanic_df[which(titanic_df$Age <=5 & 
                    titanic_df$Survived ==1),])
```

## Get unique values for a column

> How many embark points were there? dataframe.unique() lists unique values of the column

```{r}
unique(titanic_df$Sex)
```

## Working with NA Values

> How many NA values in "Embarked" columns

```{r}
table(titanic_df$Embarked)
```

## Drop NA: Strategy


```{r}
clean_titanic_df = na.omit(titanic_df)
dim(clean_titanic_df)
```

## Rename a Column

> Rename "Embarked" column to "onboarded"
> Rename "Sex" column to "Gender"

```{r}
head(titanic_df,2)
```

```{r}
colnames(titanic_df) = sub('Embarked', 'onboarded', colnames(titanic_df))

colnames(titanic_df) = sub('Sex', 'Gender', colnames(titanic_df))
```

```{r}
head(titanic_df,2)
```

## Apply filter criteria and Select columns based

> Select Age, Gender and Pclass of passengers who survived and where less than 5 years old

```{r}
head(titanic_df[ which(titanic_df$Age <= 5 & 
              titanic_df$Survived ==1), ][c('Age','Gender','Pclass')])
```

> Only age, sex, survived and pclass of passengers whose age are not known

```{r}
dim(titanic_df[ is.na(titanic_df$Age), ])
```

> Only age, sex, survived and pclass of passengers whose age are known

```{r}
dim(titanic_df[ !is.na(titanic_df$Age), ][c('Age','Gender','Pclass')])
new_titanic_df = (titanic_df[ !is.na(titanic_df$Age), ][c('Age','Gender','Pclass')])
head(new_titanic_df)
```


## Groupby and Aggregate

> Find the average age of passengers based on Pclass column

```{r}
tapply(new_titanic_df$Age,new_titanic_df$Pclass, mean)
```

```{r}
summary(titanic_df$Age)
```

## Groupby and Aggregate - dplyr

> Find the average age of passengers based on Pclass column

```{r}
titanic_df %>% group_by(Pclass) %>% summarize(Avg_Age = mean(Age, na.rm=TRUE))
```

The overall average age of passengers:

```{r}
mean(titanic_df$Age, na.rm = TRUE)
```

> Find the average age of passengers based on Pclass and Gender

```{r}

pclass_gender_age_df = titanic_df %>% group_by(Pclass,Gender) %>% summarize(Avg_Age = mean(Age, na.rm=TRUE))

pclass_gender_age_df
```

> How many passengers survived in each of the pclass and for each of the gender

```{r}

pclass_gender_sur_df = titanic_df %>% group_by(Pclass,Gender) %>% summarize(Survived_Count = sum(Survived, na.rm=TRUE))

pclass_gender_sur_df
```

Calculate the summary statistics i.e. count of observations, average age and standard deviation by Pclass.
```{r}
group_titanic_df = titanic_df %>% group_by(Pclass)

titanic_summary = summarise(group_titanic_df,
  count = n(),
  avg_age = mean(Age, na.rm = TRUE),
  sd_month = sd(Age, na.rm = TRUE))
```

## Merge Dataframe

> Merge pclass_gender_age_df with pclass_gender_sur_df based on pclass and Gender

https://www.infoworld.com/article/3454356/how-to-merge-data-in-r-using-r-merge-dplyr-or-datatable.html

```{r}
pclass_gender_merge_df = inner_join(pclass_gender_age_df,
                                    pclass_gender_sur_df,
                                    by = c('Pclass','Gender')
)

pclass_gender_merge_df
```

> Sort the pclass_gender_merge_df on Survived_Count column

```{r}
pclass_gender_merge_df %>% arrange(desc(Survived_Count))
```

Save the merged dataframe
```{r}
#write.csv(pclass_gender_merge_df, "merged_df.csv")
```

### Exercise: 

Using `pclass_gender_merge_df` do a visualization of Age and Survived w.r.t Pclass and Sex. What kind of charts will be meaningful here?

```{r}
# Write your code here.
```

## Creating Feature set (X) and Outcome Variable (Y)

```{r}
get_predictors = function(df) {
  predictors = colnames(df)[!colnames(df) %in% c('Survived')]
  return (predictors)
}

get_response = function(df) {
  response = colnames(df)[!colnames(df) %in% predictors]
  return (response)
}
```

```{r}
predictors = get_predictors(titanic_df)
predictors
response = get_response(titanic_df)
response
```

## Get the categorical and numeric features

Create a list of categorical and numeric features in a dataset:

```{r}
cat_feat_list = list(colnames(titanic_df[predictors][sapply(titanic_df[predictors], is.factor)]))

typeof(cat_feat_list)

class(cat_feat_list)

cat_feat_list = unlist(cat_feat_list)

typeof(cat_feat_list)
cat_feat_list
```


```{r}
num_feat_list = (colnames(titanic_df[predictors][sapply(titanic_df[predictors], is.numeric)]))

typeof(num_feat_list)
class(num_feat_list)
num_feat_list
```

## User defined functions

Create user defined functions

> Define a function to find NA values and unique labels in any column

```{r}
find_na = function(name){
  return(table(titanic_df[,name]))
}
```

```{r}
find_unique = function(name){
  return(unique(titanic_df[,name]))
}
```

```{r}
find_na('Gender')
find_unique('Gender')
```

Define a function to perform cross tabulation

```{r}
cross_tab = function(x,y){
  return (table(titanic_df[,x],titanic_df[,y]))
}
```

```{r}
df = cross_tab('Gender','Pclass')
df
```

## Exercise 1: 

1. Use the user defined function `find_na` and `find_unique` to get the value count and unique labels of all the categorical features in the titanic dataframe

```{r}
cat_feat_list
```

```{r}
cat_feat_list = append(cat_feat_list, "Survived")
cat_feat_list = cat_feat_list[!cat_feat_list %in% c('Name','Ticket','Cabin')]
cat_feat_list
```

```{r}
sapply(cat_feat_list,find_na)
sapply(cat_feat_list, find_unique)
```

## Exercise 2:

2. Use the user defined function `cross_tab` to cross tabulate every pair of categorical features. 

Dataframe for every pair of categorical feature is as follows

```{r}
df = expand.grid(cat_feat_list,cat_feat_list,stringsAsFactors = FALSE)
df
```

Pivot of the unqiue combination of categorical feature is as follows:

```{r}

for (i in 1:nrow(df)){
  if(df[i,"Var1"] != df[i,"Var2"]){
    cat("Pivot of", df[i,"Var1"]," with ", df[i,"Var2"], "\n")
    print(cross_tab(df[i,"Var1"],df[i,"Var2"]))
  }
}
```

## Exercise 3: DIY

3. Change the cross_tab function to generate pivots by nomalized values of rows (i.e. percentage by row sum)and columns (i.e percentage by column sum).

## Thank You